Quantum H theorem, variational entropy, and relaxation processes

### 2020-12-28 tex/quantum.tex
Rewording: start of sec. 3.2

Consider a diluted gas of $N$ non-interacting quantum particles (either bosons or fermions),
contained by a vessel of volume
$V$, at some temperature $T$, and total energy E. Starting from the Boltzmann definition
of entropy, the quantum $H$ functional is
%
\begin{equation}
   H_T=-\ln G,
\end{equation}
%
where $G$ is the total number of possible quantum states of the gas that satisfy the
above conditions \cite{bib:tolman}.
The quantum $H$-theorem can be demostrated as follows.  $G$ can be divided into groups of neighboring states,
$g_k$, and a certain occupation numbers, $n_k$, can be associated with each of these
groups. Thus, the above functional takes the form:
%
\begin{equation}\label{eq:quantumh}
    H_T=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, 
\end{equation}
%
where the upper and lower signs are for bosons and fermions, respectively.
Thus the time derivative of Eq.~(\ref{eq:quantumh}) is
%
\begin{equation}\label{eq:quantumdHdt}
\frac{dH}{dt}=\sum_{\kappa}\left[\ln n_{\kappa}-\ln\left(g_{\kappa}\pm n_{\kappa}\right)\right]
\frac{dn_{\kappa}}{dt}.
\end{equation}
%

@@

\subsection{Out of equilibrium, non-homogeneous distributions for degenerated systems}

We begin by defining the form of the functional
%
\begin{eqnarray}
    \Ss (t)&=&\sum_{M=1}^{K} \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ]   \delta V_M \delta \epsilon_n \label{entropy}.
\end{eqnarray}
%
This functional will be named \textit{variational entropy}.
This functional describes, as in the classical case, a system with a global
energy $E$, a global number of free
quantum particles (which may be fermions for upper sing or bosons for lower
sing) $N$, global volume $V$, and a temperature $T$.
Besides, the variational entropy should correspond to a density of entropy when
the system arrives at the equilibrium state. The
global system, in turn, consists of a set of $K$ small cells with a constant
volume $\delta V_M$, whose local homogeneous distribution functions is
$f_{Mn}(\epsilon_{n},t)$. The distribution function will depend on the position
of the cells, the energy per cell $\epsilon_{n}$, and time $t$. Also, the
energy spectra will be divided into $\delta \epsilon_n$ volumes. We will use
the uppercase Latin index to denote cell numbers, and unless stated otherwise,
$M = 1, 2 . . . , K$. Without loss of generality, we choose these cells to have
local identical volumes (\textit{i.e.} $\delta V_M = V/K, \forall M$). Each
cell has its local chemical potential, $\mu_M$, local
variational functional, $\Sm$, local number of particles, $\mathcal{N}_M$, and
local energy, $\mathcal{E}_M$. In terms of the local distribution
function and local energy spectra, $\{\epsilon_{n}\}$, the last three
properties are given by:
%
\begin{eqnarray}
    \Sm &=&  \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ] \de V_M \delta \epsilon_n \label{entropycell},\\
    {\mathcal{N}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n} ,t) \de V_M \delta \epsilon_n, \nonumber \\
{\mathcal{E}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n} \de V_M \delta \epsilon_n.
\end{eqnarray}
%
When the system is in equilibrium, the local number of particles and the local
energy don't depend on the cell number, this is
%
\begin{equation}
   {\mathcal{N}}_M=\mathcal{N}\equiv \bar{\mathcal{N}}; \ \ \ \  {\mathcal{E}}_M=\mathcal{E}\equiv \bar{\mathcal{E}}.
\end{equation}
%
The particles in the system are considered to be free, \textit{i.e.} the
particles do not interact with each other.
Consequently, we can assume that the available quantum states do not depend on
the cell features,
therefore the energy spectrum, denoted by $\epsilon_n$, is the same for all
cells, \textit{i.e.} $\epsilon_{Mn} = \epsilon_n, \forall M$. Besides, the
system is subjected to the microcanonical restrictions
%
\begin{eqnarray}
    &&\sum_{M=1}^{K} \left[ \sum_{n}f_{Mn}(\epsilon_{n} ,t)\de \epsilon_n\right] \de V_M=\sum_{M=1}^{K} {\mathcal{N}}_{M} \de V_M=N; \nonumber \\
    &&\sum_{M=1}^{K}\left[ \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n}\delta \epsilon_n\right] \de V_M=\sum_{M=1}^{K} {\mathcal{E}}_M \delta V_M=E, \label{restriccions1}
\end{eqnarray}
%
where $N$ and $E$ are the total particle number and the total energy of the
system respectively.

Maximizing the variational entropy $\Ss$ using the Lagrange multipliers method,
we obtain
%
\begin{eqnarray}
&&\ln \left(\frac{1\mp f_{Mn}(\epsilon_{n},t)}{f_{Mn}(\epsilon_{n},t)} \right)=-\alpha_M(t)-\beta_M(t) \epsilon_{n}, \label{relation}\\ &&\Rightarrow \ \ f_{Mn}(\epsilon_{n},t)=\frac{1}{e^{-\alpha_M(t)-\beta_M(t) \epsilon_{n}}\pm 1} \equiv \bar{f}_{Mn}(\epsilon_{n},t) \label{distributionequilibrium}.
\end{eqnarray}
%
where $\alpha_M(t), \beta_M(t)$ are the Lagrange multipliers. We can obtain
from the variational procedure that
%
\begin{equation}
    \alpha_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{N}_M}, \ \ \ \beta_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{E}_M},\label{multipliers}
\end{equation}
%
where we used that the $H$-functional in equilibrium is proportional to the
entropy of the system.
In other hand, in the expression (\ref{multipliers}), we expect that the
variation of the logarithm with the number of particles by cells is indifferent
from a particular cell. With this, we conclude 
%
\begin{equation}
    \alpha_M=\alpha; \ \ \ \ \beta_M=\beta.
\end{equation}
%
In equilibrium, the energy and the Lagrange multiplier do not depend on the
cell number and time, therefore
%
\begin{equation}
    \bar f_{Mn}(\epsilon_{n},t)=\bar f_n(\epsilon_{n}) =\frac{1}{e^{-\alpha-\beta \epsilon_n}\pm 1}.
\end{equation}
%
If one compares those multipliers with the mean energy and the mean particle
number from the statistical mechanic's results, we obtain
%
\begin{equation}
    \alpha=\frac{\mu}{kT}\equiv \bar{\alpha}; \ \ \ \ \beta=-\frac{1}{kT}\equiv \bar{\beta},
\end{equation}
%
and finally
%
\begin{eqnarray}
    \bar{f}_{n}(\epsilon_{n})&=&\frac{1}{e^{(\frac{{\epsilon_n}-\bar{\mu}}{kT})}\pm 1}\equiv \bar{f}_{n}.
\end{eqnarray}
%
We proved, when the system is in equilibrium, the Lagrange multipliers don't
depend on the cell number. In other words, Lagrange multipliers are homogeneous
in the system.

We can calculate the entropy from an ideal system, in other words, if one
substitute (\ref{distributionequilibrium}) to (\ref{entropy}) we obtain
%
\begin{eqnarray}
      \Ss(t)&=&\sum_{M=1}^{K} \sum_n  \left[\left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right)\ln \left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}} \right) \right]\nonumber \\
      &&\pm  \ln \left[\prod_{M=1}^{K} \prod_{n}\left(1 \mp \frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right) \right] \de V_M \delta \epsilon_n\label{H-entropy}.
  \end{eqnarray}
%
This result allows us to calculate the entropy of a quantum ideal gases. This
entropy could be used as a thermodynamic variable to describe those systems.

%\subsection{The quantum analogous of the Boltzmann's $H$-theorem}
We rewrite the \textit{variational entropy} (\ref{entropy}) as
%
\begin{equation}
    \Ss(t)=\sum_{M=1}^{K} \sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right] \de V_M \delta \epsilon_n \label{entropy2},
\end{equation}
%
where, for simplicity, we have defined $f(\epsilon_{n},t)\equiv f_{Mn}$.

@@

If we perform the derivative of the variational entropy with respect to time,
we obtain
%
\begin{equation}
   \frac{d \Ss (t)}{dt}= \sum_n \sum_{M=1}^{K} \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M \delta \epsilon_n.\label{deltaH}
\end{equation}
%
Using the first-order approximation (\ref{firstorder}), we get
%
\begin{eqnarray}
    \frac{d\Ss (t)}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_{n}\ln \left[ \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})} \right]\dot{g}_{nM} \de V_M\delta \epsilon_n \nonumber \\
    &=&\sum_n \sum_{M=1}^{K} \bar{f}_n \left \{ \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}-\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}  \right \}\de V_M \delta \epsilon_n.\nonumber \\
    \label{cambioH1}
\end{eqnarray}
%
It's necessary to remark that, in the case of fermions
%
\begin{eqnarray}
   1-\bar f_n -\bar f_n g_{nM}>0 \ \ \Rightarrow \ \ \frac{1}{\bar f_n}>1+g_{nM}. \label{fermionrestriction}
\end{eqnarray}
%
The previous expression establishes that the value of $g_{nM}$ is determined by
$\bar f_n$. Furthermore, whether $\bar f_{n}=1$, this is, all lower energy
levels are occupied, then the system will not have inhomogeneities because of
the exclusion principle, and as a consequence of this,  $g_{nM}=0$. This result
corresponds to the system in the zero temperature condition. We can observe
that in this condition, (\ref{fermionrestriction}) is violated. Due to this,
the zero temperature condition is excluded.

We can observe in the case of bosons, if $\bar{f}_n \gg \bar{f}_n |g_{nM}|$
then $1+\bar{f}_n \gg \bar{f}_n |g_{nM}|$. In fermions, in the case of a
non-extremely degenerated condition, it is true that $1-\bar{f}_n \gg \bar{f}_n
|g_{nM}|$. With those relations, we can approximate the logarithm functions in
(\ref{cambioH1}) to their first-order Taylor series around $\bar f_n g_{nM}=0$
%
\begin{equation}
    \ln [\bar{f}_n+\bar{f}_n g_{nM}] \approx \ln [\bar{f}_n]+ g_{nM}; \ \ \ \ \ln[1\mp\bar{f}_n\mp\bar{f}_n g_{nM}] \approx \ln[1\mp\bar{f}_n]\mp\frac{\bar{f}_n}{1\mp\bar{f}_{n}} g_{nM}. \label{lnapproximation}
\end{equation}
%
With the previous approximation, (\ref{cambioH1}) will be
%
\begin{eqnarray}
    \frac{d\Ss}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_n\left \{ (\ln \bar{f}_n+ g_{nM})\dot{g}_{nM}\right\} \de V_M \delta \epsilon_n \nonumber \\
    &&-\sum_{n}\sum_{M=1}^{K}\bar f_n\left\{ \left( \ln[1\mp\bar{f}_n]\mp \left[\frac{\bar{f}_n}{1\mp\bar{f}_n} \right] g_{nM}\right)\dot{g}_{nM} \right \}\de V_M \delta \epsilon_n.\label{cambioH2}
\end{eqnarray}
%
Making use of the expression (\ref{relation}),
(\ref{cambioH2}) casts into
%
\begin{eqnarray}
    \frac{d\Ss}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_n\left \{ (\bar{\alpha}+\bar{\beta}{\epsilon}_n)\dot{g}_{nM}+ g_{nM}\left(1\pm e^{\bar{\alpha}+\bar{\beta}{\epsilon}_n}\right)\dot{g}_{nM} \right \} \de V_M \delta \epsilon_n. \nonumber \\
    \label{cambioH3}
\end{eqnarray}
%
On the other hand, we obtain from the restrictions the following expressions
%
\begin{eqnarray}
    &&\sum_n \bar{f}_n g_{nM} \delta \epsilon_n=\Delta_M(t) \ \  \Rightarrow \ \  \sum_n \bar{f}_n \dot{g}_{nM} \delta \epsilon_n=\dot{\Delta}_M(t), \nonumber \\
    &&\sum_n  \bar{f}_n g_{nM}\epsilon_n \delta \epsilon_n=\delta_M(t) \ \  \Rightarrow \ \  \sum_n \bar{f}_n \dot{g}_{nM}\epsilon_n \delta \epsilon_n=\dot{\delta}_M(t)
\end{eqnarray}
%
and as a consequence of $\sum_{M=1}^{K} \Delta_M(t) \de V_M  =\sum_{M=1}^{K} \delta_M(t) \de V_M =0$, we find
%
\begin{equation}
    \sum_{M=1}^{K} \dot{\Delta}_M(t) \de V_M =\sum_{M=1}^{K} \dot{\delta}_{M}(t) \de V_M=0.
\end{equation}
%
Substituting the previous expression to (\ref{cambioH3}) we obtain
%
\begin{eqnarray}
   \frac{d\Ss}{dt}&=&  \sum_n e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\sum_M  g_{nM}\dot{g}_{nM} \de V_M\delta \epsilon_n. \nonumber \\ \label{cambioH4}
\end{eqnarray}
%
Now, the summation over $M$ will be expressed in two summations, such as in the
classical case
%
\begin{equation}
    \frac{d\Ss}{dt}=\sum_n  e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\left(\sum_J ^{L} g^{+}_{nJ}\dot{g}^{+}_{nJ}\de V_J \delta \epsilon_n+\sum^{P}_J  g^{-}_{nJ}\dot{g}^{-}_{nJ} \de V_J\delta \epsilon_n \right), \label{cambioH5}
\end{equation}
%
where $L+P=K$. Besides, $\dot{g}^{+}_{J}$ represents the change on the
deviation on cells that have an excess of particles or energy while
$\dot{g}^{-}_{J}$  represents the change on the deviations on cells that have
missing particles or energy.
On the other hand, $g^{+}_{J}$  represents the value of the deviation on cells
that have an excess of particles or energy. In contrast, $g^{-}_{J}$ represents
the value of the deviation on cells that have missing particles or energy.

@@

and consequently, (\ref{cambioH5}) obtains the following form
%
\begin{equation}
    \frac{d\Ss}{dt}=-\sum_n  e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\left(\sum_J ^{L} |g^{+}_{nJ}||\dot{g}^{+}_{nJ}|\de V_J \delta \epsilon_n+\sum^{P}_J  |g^{-}_{nJ}||\dot{g}^{-}_{nJ}| \de V_J \delta \epsilon_n \right), \label{cambioH6}
\end{equation}
%
and due to $e^{\bar{\alpha}+\bar{\beta}\epsilon_n}$ is always positive, then
$\frac{d\Ss}{dt}<0$. With this, we proved that any quantum ideal gas perturbed
(out of equilibrium but not so far from it), evolves such that
$\frac{d\Ss}{dt}<0$.

When the system is in equilibrium, $g_{nM}=0$ and in consequence of that, $\dot
g_{nM}=0$ and finally from (\ref{cambioH1}) $\frac{d\Ss}{dt}=0$. Then, we can
say that any ideal quantum gas perturbed always evolves to the equilibrium
state, such as the classical $H$-theorem. Then, we obtained the quantum version
of the Boltzmann's $H$-theorem including inhomogeneities in the system.

>>>>

Consider a diluted gas of $N$ non-interacting quantum particles (either bosons or fermions),
contained by a vessel of volume
$V$, at some temperature $T$, and total energy E. Starting from the Boltzmann definition
of entropy, the quantum $H$ functional is
%
\begin{equation}
   H_T=-\ln G,
\end{equation}
%
where $G$ is the total number of possible quantum states of the gas that satisfy the
above conditions \cite{bib:tolman}.
The quantum $H$-theorem can be demostrated as follows.  $G$ can be divided into groups of neighboring states,
$g_k$, and a certain occupation numbers, $n_k$, can be associated with each of these
groups. Thus, the above functional takes the form:
%
\begin{equation}\label{eq:quantumh}
    H_T=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, 
\end{equation}
%
where the upper and lower signs are for bosons and fermions, respectively.
Thus the time derivative of Eq.~(\ref{eq:quantumh}) is
%
\begin{equation}\label{eq:quantumdHdt}
\frac{dH_T}{dt}=\sum_{\kappa}\left[\ln n_{\kappa}-\ln\left(g_{\kappa}\pm n_{\kappa}\right)\right]
\frac{dn_{\kappa}}{dt}.
\end{equation}
%

@@

\subsection{Out of equilibrium, non-homogeneous quantum systems}

Consider a diluted gas enclosed by a perfectly isolated vessel of volume $V$, with
total energy $E$, and total number of quantum particles $N$, which can be fermions or
bosons. For our purposes, the volume $V$ is divided into $K$ small cells, each of which
has a constant volume $\delta V_M=V/K$, a temperature $T_M$, a local energy $\epsilon_M$,
a local number of particles $\mathcal{N}_M$, and a local set of distribution functions,
$\{f_{Mn}(t)\}$. Each $f_{Mn}$ $\big(\equiv f(\vec r_M,\epsilon_{n},t)\big)$ represent the number
of particles contained in the $M$-th cell that occupy the energy level $\epsilon_n$.
We propose the following functional as an alternative $H$-functional
for quantum non-homogeneous diluted gases:
%
\begin{eqnarray}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ]   \delta V_M \delta \epsilon_n \label{entropy}.
\end{eqnarray}
%
Here, the upper and lower sign refer to bosons and fermions respectively.
\color{blue}\commr{Also, the energy spectrum will be divided into $\delta \epsilon_n$ volumes}. We will use
the uppercase Latin index to denote cell numbers, and unless stated otherwise,
$M = 1, 2 . . . , K$. Without loss of generality, we choose these cells to have
local identical volumes (\textit{i.e.} $\delta V_M = V/K, \forall M$). Each
cell has its local chemical potential, $\mu_M$, local
variational functional, $\mathcal{H}_M$, local number of particles, $\mathcal{N}_M$, and
local energy, $\mathcal{E}_M$. In terms of the local distribution
function and local energy spectra, $\{\epsilon_{n}\}$, the last three
properties are given by:
%
\begin{eqnarray}
    \mathcal{H}_M &=&  \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ] \de V_M \delta \epsilon_n \label{entropycell},\\
    {\mathcal{N}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n} ,t) \de V_M \delta \epsilon_n, \nonumber \\
{\mathcal{E}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n} \de V_M \delta \epsilon_n.
\end{eqnarray}
%
When the system is in equilibrium, the local number of particles and the local
energy don't depend on the cell number, this is
%
\begin{equation}
   {\mathcal{N}}_M=\mathcal{N}\equiv \bar{\mathcal{N}}; \ \ \ \  {\mathcal{E}}_M=\mathcal{E}\equiv \bar{\mathcal{E}}.
\end{equation}
%
The particles in the system are considered to be free, \textit{i.e.} the
particles do not interact with each other.
Consequently, we can assume that the available quantum states do not depend on
the cell features,
therefore the energy spectrum, denoted by $\epsilon_n$, is the same for all
cells, \textit{i.e.} $\epsilon_{Mn} = \epsilon_n, \forall M$. Besides, the
system is subjected to the microcanonical restrictions
%
\begin{eqnarray}
    &&\sum_{M=1}^{K} \left[ \sum_{n}f_{Mn}(\epsilon_{n} ,t)\de \epsilon_n\right] \de V_M=\sum_{M=1}^{K} {\mathcal{N}}_{M} \de V_M=N; \nonumber \\
    &&\sum_{M=1}^{K}\left[ \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n}\delta \epsilon_n\right] \de V_M=\sum_{M=1}^{K} {\mathcal{E}}_M \delta V_M=E, \label{restriccions1}
\end{eqnarray}
%
where $N$ and $E$ are the total particle number and the total energy of the
system respectively.

Maximizing the variational entropy $\mathcal{H}$ using the Lagrange multipliers method,
we obtain
%
\begin{eqnarray}
&&\ln \left(\frac{1\mp f_{Mn}(\epsilon_{n},t)}{f_{Mn}(\epsilon_{n},t)} \right)=-\alpha_M(t)-\beta_M(t) \epsilon_{n}, \label{relation}\\ &&\Rightarrow \ \ f_{Mn}(\epsilon_{n},t)=\frac{1}{e^{-\alpha_M(t)-\beta_M(t) \epsilon_{n}}\pm 1} \equiv \bar{f}_{Mn}(\epsilon_{n},t) \label{distributionequilibrium}.
\end{eqnarray}
%
where $\alpha_M(t), \beta_M(t)$ are the Lagrange multipliers. We can obtain
from the variational procedure that
%
\begin{equation}
    \alpha_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{N}_M}, \ \ \ \beta_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{E}_M},\label{multipliers}
\end{equation}
%
where we used that the $H$-functional in equilibrium is proportional to the
entropy of the system.
In other hand, in the expression (\ref{multipliers}), we expect that the
variation of the logarithm with the number of particles by cells is indifferent
from a particular cell. With this, we conclude 
%
\begin{equation}
    \alpha_M=\alpha; \ \ \ \ \beta_M=\beta.
\end{equation}
%
In equilibrium, the energy and the Lagrange multiplier do not depend on the
cell number and time, therefore
%
\begin{equation}
    \bar f_{Mn}(\epsilon_{n},t)=\bar f_n(\epsilon_{n}) =\frac{1}{e^{-\alpha-\beta \epsilon_n}\pm 1}.
\end{equation}
%
If one compares those multipliers with the mean energy and the mean particle
number from the statistical mechanic's results, we obtain
%
\begin{equation}
    \alpha=\frac{\mu}{kT}\equiv \bar{\alpha}; \ \ \ \ \beta=-\frac{1}{kT}\equiv \bar{\beta},
\end{equation}
%
and finally
%
\begin{eqnarray}
    \bar{f}_{n}(\epsilon_{n})&=&\frac{1}{e^{(\frac{{\epsilon_n}-\bar{\mu}}{kT})}\pm 1}\equiv \bar{f}_{n}.
\end{eqnarray}
%
We proved, when the system is in equilibrium, the Lagrange multipliers don't
depend on the cell number. In other words, Lagrange multipliers are homogeneous
in the system.

We can calculate the entropy from an ideal system, in other words, if one
substitute (\ref{distributionequilibrium}) to (\ref{entropy}) we obtain
%
\begin{eqnarray}
      \mathcal{H}(t)&=&\sum_{M=1}^{K} \sum_n  \left[\left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right)\ln \left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}} \right) \right]\nonumber \\
      &&\pm  \ln \left[\prod_{M=1}^{K} \prod_{n}\left(1 \mp \frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right) \right] \de V_M \delta \epsilon_n\label{H-entropy}.
  \end{eqnarray}
%
This result allows us to calculate the entropy of a quantum ideal gases. This
entropy could be used as a thermodynamic variable to describe those systems.

%\subsection{The quantum analogous of the Boltzmann's $H$-theorem}
We rewrite the \textit{variational entropy} (\ref{entropy}) as
%
\begin{equation}
    \mathcal{H}(t)=\sum_{M=1}^{K} \sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right] \de V_M \delta \epsilon_n \label{entropy2},
\end{equation}
%
where, for simplicity, we have defined $f(\epsilon_{n},t)\equiv f_{Mn}$.

@@

If we perform the derivative of the variational entropy with respect to time,
we obtain
%
\begin{equation}
   \frac{d \mathcal{H} (t)}{dt}= \sum_n \sum_{M=1}^{K} \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M \delta \epsilon_n.\label{deltaH}
\end{equation}
%
Using the first-order approximation (\ref{firstorder}), we get
%
\begin{eqnarray}
    \frac{d\mathcal{H} (t)}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_{n}\ln \left[ \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})} \right]\dot{g}_{nM} \de V_M\delta \epsilon_n \nonumber \\
    &=&\sum_n \sum_{M=1}^{K} \bar{f}_n \left \{ \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}-\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}  \right \}\de V_M \delta \epsilon_n.\nonumber \\
    \label{cambioH1}
\end{eqnarray}
%
It's necessary to remark that, in the case of fermions
%
\begin{eqnarray}
   1-\bar f_n -\bar f_n g_{nM}>0 \ \ \Rightarrow \ \ \frac{1}{\bar f_n}>1+g_{nM}. \label{fermionrestriction}
\end{eqnarray}
%
The previous expression establishes that the value of $g_{nM}$ is determined by
$\bar f_n$. Furthermore, whether $\bar f_{n}=1$, this is, all lower energy
levels are occupied, then the system will not have inhomogeneities because of
the exclusion principle, and as a consequence of this,  $g_{nM}=0$. This result
corresponds to the system in the zero temperature condition. We can observe
that in this condition, (\ref{fermionrestriction}) is violated. Due to this,
the zero temperature condition is excluded.

We can observe in the case of bosons, if $\bar{f}_n \gg \bar{f}_n |g_{nM}|$
then $1+\bar{f}_n \gg \bar{f}_n |g_{nM}|$. In fermions, in the case of a
non-extremely degenerated condition, it is true that $1-\bar{f}_n \gg \bar{f}_n
|g_{nM}|$. With those relations, we can approximate the logarithm functions in
(\ref{cambioH1}) to their first-order Taylor series around $\bar f_n g_{nM}=0$
%
\begin{equation}
    \ln [\bar{f}_n+\bar{f}_n g_{nM}] \approx \ln [\bar{f}_n]+ g_{nM}; \ \ \ \ \ln[1\mp\bar{f}_n\mp\bar{f}_n g_{nM}] \approx \ln[1\mp\bar{f}_n]\mp\frac{\bar{f}_n}{1\mp\bar{f}_{n}} g_{nM}. \label{lnapproximation}
\end{equation}
%
With the previous approximation, (\ref{cambioH1}) will be
%
\begin{eqnarray}
    \frac{d\mathcal{H}}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_n\left \{ (\ln \bar{f}_n+ g_{nM})\dot{g}_{nM}\right\} \de V_M \delta \epsilon_n \nonumber \\
    &&-\sum_{n}\sum_{M=1}^{K}\bar f_n\left\{ \left( \ln[1\mp\bar{f}_n]\mp \left[\frac{\bar{f}_n}{1\mp\bar{f}_n} \right] g_{nM}\right)\dot{g}_{nM} \right \}\de V_M \delta \epsilon_n.\label{cambioH2}
\end{eqnarray}
%
Making use of the expression (\ref{relation}),
(\ref{cambioH2}) casts into
%
\begin{eqnarray}
    \frac{d\mathcal{H}}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_n\left \{ (\bar{\alpha}+\bar{\beta}{\epsilon}_n)\dot{g}_{nM}+ g_{nM}\left(1\pm e^{\bar{\alpha}+\bar{\beta}{\epsilon}_n}\right)\dot{g}_{nM} \right \} \de V_M \delta \epsilon_n. \nonumber \\
    \label{cambioH3}
\end{eqnarray}
%
On the other hand, we obtain from the restrictions the following expressions
%
\begin{eqnarray}
    &&\sum_n \bar{f}_n g_{nM} \delta \epsilon_n=\Delta_M(t) \ \  \Rightarrow \ \  \sum_n \bar{f}_n \dot{g}_{nM} \delta \epsilon_n=\dot{\Delta}_M(t), \nonumber \\
    &&\sum_n  \bar{f}_n g_{nM}\epsilon_n \delta \epsilon_n=\delta_M(t) \ \  \Rightarrow \ \  \sum_n \bar{f}_n \dot{g}_{nM}\epsilon_n \delta \epsilon_n=\dot{\delta}_M(t)
\end{eqnarray}
%
and as a consequence of $\sum_{M=1}^{K} \Delta_M(t) \de V_M  =\sum_{M=1}^{K} \delta_M(t) \de V_M =0$, we find
%
\begin{equation}
    \sum_{M=1}^{K} \dot{\Delta}_M(t) \de V_M =\sum_{M=1}^{K} \dot{\delta}_{M}(t) \de V_M=0.
\end{equation}
%
Substituting the previous expression to (\ref{cambioH3}) we obtain
%
\begin{eqnarray}
   \frac{d\mathcal{H}}{dt}&=&  \sum_n e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\sum_M  g_{nM}\dot{g}_{nM} \de V_M\delta \epsilon_n. \nonumber \\ \label{cambioH4}
\end{eqnarray}
%
Now, the summation over $M$ will be expressed in two summations, such as in the
classical case
%
\begin{equation}
    \frac{d\mathcal{H}}{dt}=\sum_n  e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\left(\sum_J ^{L} g^{+}_{nJ}\dot{g}^{+}_{nJ}\de V_J \delta \epsilon_n+\sum^{P}_J  g^{-}_{nJ}\dot{g}^{-}_{nJ} \de V_J\delta \epsilon_n \right), \label{cambioH5}
\end{equation}
%
where $L+P=K$. Besides, $\dot{g}^{+}_{J}$ represents the change on the
deviation on cells that have an excess of particles or energy while
$\dot{g}^{-}_{J}$  represents the change on the deviations on cells that have
missing particles or energy.
On the other hand, $g^{+}_{J}$  represents the value of the deviation on cells
that have an excess of particles or energy. In contrast, $g^{-}_{J}$ represents
the value of the deviation on cells that have missing particles or energy.

@@

and consequently, (\ref{cambioH5}) obtains the following form
%
\begin{equation}
    \frac{d\mathcal{H}}{dt}=-\sum_n  e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\left(\sum_J ^{L} |g^{+}_{nJ}||\dot{g}^{+}_{nJ}|\de V_J \delta \epsilon_n+\sum^{P}_J  |g^{-}_{nJ}||\dot{g}^{-}_{nJ}| \de V_J \delta \epsilon_n \right), \label{cambioH6}
\end{equation}
%
and due to $e^{\bar{\alpha}+\bar{\beta}\epsilon_n}$ is always positive, then
$\frac{d\mathcal{H}}{dt}<0$. With this, we proved that any quantum ideal gas perturbed
(out of equilibrium but not so far from it), evolves such that
$\frac{d\mathcal{H}}{dt}<0$.

When the system is in equilibrium, $g_{nM}=0$ and in consequence of that, $\dot
g_{nM}=0$ and finally from (\ref{cambioH1}) $\frac{d\mathcal{H}}{dt}=0$. Then, we can
say that any ideal quantum gas perturbed always evolves to the equilibrium
state, such as the classical $H$-theorem. Then, we obtained the quantum version
of the Boltzmann's $H$-theorem including inhomogeneities in the system.
